## Chapter 6: Analysis of Longitudinal Data.
In this last chapter, we will perform the analyses of the MABS at the book's chapters 8-9. As for the chapter 8, we will utilize the ```RATS``` dataset, whilst for the chapter 9 we will use the ```BPRS``` dataset. Both dataset were transformed in long data format in the 6.1.1.Data Wrangling Exercise available [here](https://github.com/015129374/IODS-project/blob/master/Data/meet_and_repeat.R).

### 6.2.1. Analysis of Longitudinal Data I: Graphical Displays and Summary Measure Approach with the RATS dataset.
The ```RATS``` dataset comes from a nutrition study by Crowder and Hand (1990), on three groups of rats which were fed differently and had their weight recorded over a nine-week period. The main question is whether the growth profiles differ.
After reading the table with the function ```read.table()```, we factorize the variables ```"ID"``` and ```"Group"```, and we add the variable ```"Time"```, now we have the ```RATS``` in the long format ```RATSL```:

```{r, message = FALSE}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep ="\t", header = TRUE)
library(dplyr); library(tidyr)
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
RATSL <- gather(RATS, key = WD, value = Weight, -ID, - Group) %>% mutate(Time = as.integer(substr(WD,3,4)))
```

Then we can see both the structure and the summary of long-format dataset ```RATSL```:
```{r, echo = FALSE, message = FALSE}
str(RATSL)
summary(RATSL)
```

We can also see the ```head()``` and ```tail()``` of the ```RATSL``` dataset:
```{r, echo = FALSE, message = FALSE}
head(RATSL)
tail(RATSL)
```

#### Graphical Displays of Longitudinal Data.  
The graphical display of longitudinal data is helpful in showing as much as raw data as possible and in unveiling different patterns such as aggregate, cross-sectional and longitudinal. It also allows for a simplier detection of unusual individuals or observations. Here we proceed in produce the first plot for all the rats in ```RATSL``` divided by the three groups:
```{r}
library(tidyr); library(dplyr); library(ggplot2)
p1 <-ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
p6
```

###### __Figure 8.1__: Individual "Weight" recordings by groups for the RATS data.

From this figure we can see that the weight has increased over the period of time for the large majority of the rats, more steeply in the ```"Group 2"``` and ```"Group 3"```. As for the ```"Group 1"``` the density of the lines makes the figure difficult to read but the weight seem to have increased in a less pronounced fashion. Each group also present an outlier. From the figure it emerges also that rats with higher values at the beginning tend to have higher values throughout the study,this is the so-called phenomenon of tracking. This was the visualization of the data as they were, now we will proceed to standardize the scores:
```{r}
RATSL <- RATSL %>%
 group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()
glimpse(RATSL)
```

After the standardization we draw the plot that visualizes the standardized ```"Weight"```:

```{r}
p1 <-ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(name = "standardized Weight")
p6
```

###### __Figure 8.2__: Individual "Weight" recordings by groups for RATS data after standardization.  
The phenomenon of tracking is more visibile from this graphic. Still, this type of visualization is not very clear. Let's then proceed to some alternative visualization options.

#### Summary Measure Analysis of Longitudinal Data.  
Within the summary measure methods there are better visualization options. The first consists in the visualization of the average weight for each group together with the indication of the variations of the observation at each timepoint. We will see this option in the __Figure 8.3__. A plotting alternative is a graph side-by-side boxplot of the observations at each time point, as in the __Figure 8.4__. 

Here we have the amount of ```Time```, baseline (```"WD1"```) included:
```{r}
n <- RATSL$Time %>% unique() %>% length()
```

The following function produces a summary data:
```{r}
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se=sd(Weight)/sqrt(n)) %>%
  ungroup()
glimpse(RATSS)
RATSS
```

which is then used to obtain the plot for mean ```Weight```:
```{r}
p1 <- ggplot(RATSS, aes (x = Time, y = mean, linetype = Group, shape = Group))
p2 <- p1 + geom_line() + scale_linetype_manual(values = c(3:5))
p3 <- p2 + geom_point(size=3) + scale_shape_manual(values = c(3:5))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + theme(legend.position = "right")
p7 <- p6 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
p7
```

###### __Figure 8.3__: Mean "Weight" by groups for the RATS data.
Apart from the different values in ```Weight```, we can see that especially the ```Group 2``` and the ```Group 3```, present similar dynamics in weight gaining. The ```Group 1```, has a less steep weight gain and does not show the sudden drop after ```WD40``` as in the other two groups.

Let's now proceed to visualize the side-by-side boxplot, which is created by executing the chunk below:
```{r}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + theme(legend.position = "right")
p5 <- p4 + scale_x_discrete(name = "Time")
p6 <- p5 + scale_fill_grey(start = 0.5, end = 1)
p6
```

###### __Figure 8.4__: Boxplot for the RATS data.
In here it is possible to note the outlier values for each group, of which, in the ```"Group 2"``` is present the highest weight. The outlier in the ```Group 1``` shows the lowest weight, whilst in the ```"Group 3"``` we find a serie of outliers that falls in the IQR of the "Group 2"```. This visualization option is certainly clearer than those represented by the __Figure 8.1__ and __Figure 8.2__.


#### Applying the Summary Measure Approach.
In here we make a summary data by using the mean of the ```"Time"``` variable:
```{r}
RATSF <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean = mean(Weight)) %>%
  ungroup()
glimpse(RATSF)
```

and then we create a boxplot of the measure for each group:
```{r}
library(ggplot2)
p1 <- ggplot(RATSF, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape = 23, size = 4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Time > 1")
p5
```

###### __Figure 8.5__: Boxplot of mean summary measures for the three groups for the RATS data.
From the figure we notice that the plots for the ```"Group 2"``` and the ```"Group 3"``` are rather skewed and they present more variable means. All the group present an outlier, but especially in the ```"Group 2"``` the value is so high that it can bias the result.

For this reason I remove the outliers, in a rather rudimental way, after checking the values I filter them one by one:
```{r}
RATSFS1 <- RATSF %>%
  filter(mean < 590)%>%
  filter(mean > 239)%>%
  filter(mean != 495.2)

glimpse(RATSFS1)
```

Then with the data cleared from the outliers in each class, I proceed to create the new boxplots:
```{r}
p1 <- ggplot(RATSFS1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape = 23, size = 4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Time > 1")
p5
```

###### __Figure 8.6__: Boxplot of mean summary measures for the three groups for the RATS data, without the ouliers.
Without the outliers the boxplots are much less skewed, still in the ```"Group 2"``` and ```"Group 3"``` the means have more variability than in the ```"Group 1"```. The distributions in the three groups seem to be a bit different, I am interested in ```"Group 2"``` and ```"Group 3"```. We shall then use a t-test to check this statement.

We will use the ```RATSFS1``` data without outliers leaving out the ```"Group 1"```:
```{r}
RATSFS1 <- RATSFS1 %>%
  filter(RATSFS1$Group != 1) 

glimpse(RATSFS1)
```

So now we run the ```t.test()```:

```{r}
t.test(mean ~ Group, data = RATSFS1, var.equal = TRUE)
```
The two groups show differences as the report indicates. 

I also wanted to add the baseline weight to the ```RATSLFS1``` to add a bit more accuracy. However it was not possible as the ```baseline``` resulted ```NULL```.

```{r, eval= FALSE}
baseline <- RATS$Time
RATSFS2 <- RATSFS1 %>%
  mutate(baseline)
```


This also prevented to fit the ANCOVA model:
```{r, error = FALSE}
#fit <- lm(mean ~ baseline + Group, data = RATSFS2)

#summary(fit)
#anova(fit)
```

### 6.2.2. Analysis of Longitudinal Data II: Linear Mixed Effects Models for Normal Response Variables with the BPRS data.  
The ```BPRS``` dataset comes from Davis(2002), and it is about 40 male subjects whom were randomly assigned to one of two treatment groups and each individual was rated on the *brief psychiatric rating scale (BPRS)* before the beginning of the treatment and after the conclusion of it. The scaled measures the levels of 18 symptoms from one to seven and it is used to assess the presence of schizophrenia. After reading the ```BPRS``` dataset, I factorized the variables ```"treatment"```, and ```"subject"``` and produced the long-format data ```BPRSL```.

```{r, echo=FALSE}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep =" ", header = TRUE)
library(dplyr); library(tidyr)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
BPRSL <- gather(BPRS, key = weeks, value = bprs, -treatment, -subject) %>% mutate(week = as.integer(substr(weeks, 5,5)))
```

Whose structure and ```summary()``` are the following:
```{r, echo=FALSE}
str(BPRSL)
summary(BPRSL)
```

And here are also the ```head()``` and ```tail()``` of the ```BPRSL```:
```{r, echo = FALSE}
head(BPRSL)
tail(BPRSL)
```

#### Linear Mixed Effects Methods for Repeated Measures Data.

##### Fitting the Independence Model to the BPRS Data.  

In here we produce the plot of ```"bprs"``` against ```Week``` for the two treatment groups:
```{r}
p1 <-ggplot(BPRSL, aes(x= week, y = bprs, group = treatment))
p2 <- p1 + geom_text(aes(label = subject))
p3 <- p2 + scale_x_discrete(name = "Week")
p4 <- p3 + scale_y_discrete(name = "bprs")
p5 <- p4 + theme_bw()
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

###### __Figure 9.1__: Plot of bprs against weeks for treatments of the BPRSL data.
The figure is not particularly helpful in assigning each observation to its related treatment group because it contains a lot of observations. Nevertheless, it still retain some pattern which is however hard to observe. Let's then move to fit a *linear mixed model*.

##### Fitting Linear Mixed Models to the BPRS Data.  
Now we will produce a visualization of ```"bprs"``` values for each ```"subject"```, at first we do a linear regression model to ```BPRSL``` with ```bprs``` as response variable and ```treatment``` and ```week``` as explanatory variables:
```{r}
BPRS_reg <-lm(bprs ~ weeks + treatment, data = BPRSL)
summary(BPRS_reg)
```

After this ```LRM``` we proceed to visualize the plot of the ```"treatment"``` group's```"bprs"``` over time:

```{r, fig.height = 10, fig.width = 10}
p1 <- ggplot(BPRSL, aes(x = week, y =bprs, group = subject))
p2 <- p1 + geom_line(aes(linetype = subject))
p3 <- p2 + scale_x_continuous(name = "Week")
p4 <- p3 + scale_y_continuous(name = "bprs") 
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

###### __Figure 9.2__: Plot of individual bprs over time.
The plot looks quite chaotic, but it shows the the decrease of the ```bprs``` scores over ```"week"```s. As this was about the ```subject```s,  we can do another plot with the ```"treatment"``` that would result more clear.

```{r}
library(ggplot2)
p1 <- ggplot(BPRSL, aes(x = week, y =bprs, group = treatment))
p2 <- p1 + geom_line(aes(linetype = treatment))
p3 <- p2 + scale_x_continuous(name = "Week")
p4 <- p3 + scale_y_continuous(name = "bprs") 
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

###### __Figure 9.2*bis*__: Plot of treatment groups' bprs over time.
The ```"treatment 1"``` seem to have an increased ```"bprs"``` at the end of the 8 ```"week"```s, whilst the ```"treatment 2"``` shows lower values. In both groups the central phase between 2 and 6 presents the lower ```"bprs"``` values which then raise again towards the end even though not as high as the end of the ```"week"``` 1 for the group 2.


```{r, fig.width=10, fig.height=10}
pairs(BPRS, cex = 0.8)
```

###### __Figure 9.3__: scatterplot of repeated measures in subjects bprs data.

Here we have the results from __Fitting Random Intercept Model__ with ```week``` and ```treatment``` as explanatory variables to the ```BPRS``` data:
```{r, message=FALSE}
library("lme4")
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

And in here there are the results from Fitting the Random Intercept and Slope Model, with ```week``` and ```treatment``` as explanatory variables in the ```BPRS``` data:
```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
```

This is the application of the ANOVA test:
```{r}
anova(BPRS_ref1, BPRS_ref)
```

Here are the results from Fitting the Random Intercept and Slope Model that allows for a ```treatment``` X ```week``` interaction to the ```BPRS``` data:
```{r}
BPRS_ref2 <-lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
```


We can then plot the fitted bprs for each subject:
```{r, fig.width=9, fig.height=9}
Fitted <- fitted(BPRS_ref2)
BPRSL <- BPRSL %>% mutate(Fitted)

p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_line(aes(linetype = subject))
p3 <- p2 + scale_x_continuous(name = "week")
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Observed")
graph1 <- p7

p1 <- ggplot(BPRSL, aes(x = week, y = Fitted, group = subject))
p2 <- p1 + geom_line(aes(linetype = subject))
p3 <- p2 + scale_x_continuous(name = "week")
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle ("Fitted")
graph2 <- p7

graph1; graph2
```

###### __Figure 9.4__: Fitted bprs profiles from the interaction model and observed bprs profiles.
