# Chapter 2: Regression and Model Validation.  

## 2.1.Data Analysis.  
### 2.1.1.Reading of the dataframe "learning2014" and exploration of its dimension and structure.  
First of all, I read the new dataframe with the function ```read.table()```. 
```{r echo=FALSE}
learning2014 <- read.table("./data/learning2014.txt", sep= "\t", header = TRUE)
```

```{r echo=FALSE}
learning2014$X <- NULL
```

Then I use of the function ```dim()``` to show the dimension of the dataframe that is of 166 objects and 7 variables as explained in the above-mentioned data wrangling section:
```{r echo=FALSE}
dim(learning2014)
```

By typing the function ```str()```, it shows the structure of the dataframe:
```{r echo=FALSE}
str(learning2014)
```

### 2.1.2.Graphical overview of the data and summary of their variables.  
To visualize the data I type the functions ```install.packages()``` to install the visualization packages ```ggplot2``` and ```GGally```. Then by using the function ```library()``` I open them in the project.


install.packages("ggplot2")  
install.packages("GGally")  
library(ggplot2)  


#### Visualizing and Exploring the dataframe.  
The libraries are opened:  
```{r echo=FALSE, eval=FALSE}
p1 <- ggplot(learning2014, aes(x = Attitude, y = Points, col = gender))
```

```{r echo=FALSE, eval=FALSE}
p2 <- p1 + geom_point()
```

```{r echo=FALSE, eval=FALSE}
p2
```

```{r echo=FALSE, eval=FALSE}
p3 <- p2 + geom_smooth(method = "lm")
```

```{r echo=FALSE, eval=FALSE}
p4 <- ggtitle("Student's attitude versus exam points")
```

```{r echo=FALSE, eval=FALSE}
p2 
```

```{r echo=FALSE}
library(GGally)
```

```{r echo=FALSE}
library(ggplot2)
```

By using the fast plotting function ```pairs()```, we can see a scatterplot matrix resulting from the draws of all the possible scatterplots from the columns of the dataframe. Different colors are used for males and females. 

```{r echo=FALSE}
pairs(learning2014 [-1], col = learning2014$gender)
```


This second plot matrix is more advanced, and it is made with ```ggpairs()```.
```{r echo=FALSE}
p <-ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list (combo = wrap ("facethist", bins = 20)))
```

```{r echo=FALSE}
p
```

The summary of the variables:
```{r echo=FALSE}
summary(learning2014)
```

#### Output interpretation and description, distribution of the variables and their in-between relations.  
The females are almost double of the males who present a wider age range. The summary suggest a significant correlation for surf vs deep and points vs attitude.


### 2.1.3.Multiple Regression Model.  
I have chosen three variables "attitude", "deep learning" and "surface learning", with the target variable "exam points" to fit a regression model analysis.  

```{r echo=FALSE}
library(ggplot2)
```

Drawing a plot matrix with ```ggpairs()```.  
```{r echo=FALSE}
ggpairs(learning2014, lower = list(combo = wrap("facethist", bins = 20)))
```

Fitting the regression models with three explanatory variables and running the summary:
```{r echo=FALSE}
regression_model <- lm(Points ~ Attitude + deep + surf, data = learning2014)
```

```{r echo=FALSE}
summary(regression_model)
```

#### Commentary and interpretation of the results.  
__MISSING__
#### Explanation and interpretation of the statistical test related to the model parameters.  
__MISSING__
#### 2.1.4.Explanation of the relationships between the chosen explanatory variables and the target variable. With the summary of the fitted model (interpret the model parameters). Explaination and interpretation of the multiple R squared of the model.  
The adjusted R square of 0.1876 denotes a poorly fitting function to the explanatory variables. Only attitude presents statistical significance.


### 2.1.5.Diagnostic plots  

#### Residuals _vs_ Fitted values.  
```{r echo = FALSE}
plot(lm(Points ~ Attitude + deep + surf, data = learning2014), which = 1)
```

#### Normal QQ-plot.  
```{r echo = FALSE}
plot(lm(Points ~ Attitude + deep + surf, data = learning2014), which = 2)
```

#### Residuals _vs_ Leverage.  
```{r echo = FALSE}
plot(lm(Points ~ Attitude + deep + surf, data = learning2014), which = 5)
```


#### __Explanation of the model's assumptions and interpretation of their validity on the bases of the diagnostic plots.__  
A Multiple Linear Regression Model has few assumptions:  
  a _linear relationship_ between the target variable and the explanatory variables,    usually revealed by scatterplots;  
  a _multivariate normality_, which means that the residuals are normally distributed.   The QQ-plot can reveal it;  
  the _absence of multicollinearity_, in other words, the explanatory variables are     not highly correlated to each other.  
  _homoscedasticity_: or constant variance of errors. There is a similar variance of error terms across the values of the explanatory variable. A plot of standardized residuals versus predicted values shows if the points are equally distributed across all values of the dependent variables.  
The diagnostic plots delivered the following observations:  
In the __residuals _vs_ fitted values__, the plot is utilized to check the assumption of linear relationship. An horizontal line, without distinct patterns is an indicator for a linear relationship, in this case, the red line is more or less horizontal at zero. Hence here linear relationship could be assumed.  
In the __normal QQ-plot__ the plot reveals the presence of multivariate normality. A good indication is if the residual points follow the straight dashed line. For the majority it is the case here, hence normality can also be assumed.  
In the __residuals _vs_ leverage__, the plot identifies the impact of a single observation on the model. Influential points lie at the upper or at the lower right corner in a position where they are influential against a regression line. In this case the points are on the left side of the plot, thus we can say that there is no leverage.

```{r echo=FALSE, eval=FALSE}
[Reference](http://www.sthda.com/english/articles/39-regression-model-diagnostics/161-linear-regression-assumptions-and-diagnostics-in-r-essentials/)
```

